Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2025-01-23 analysis/paper/paper.Rmd
add data statement
MS revisions in progress

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022; @eacker_web-based_2019; @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

@@

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, can improve viability projections on changing
landscapes, and can help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach will help to identify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018].

@@

@@

We focus on analyzing survival of collared females and the ratio of calves to cows in aerial
surveys of groups that include collared females [@nbckc_nbckc_2022],
because these remain the most widely available sources of demographic
information for boreal caribou populations [@johnson_science_2020].
Surveyed groups also contain uncollared animals so sample size is
typically larger for recruitment than for survival estimates. Errors
that can bias composition survey results include misidentifying young
bulls as cows or vice versa, and failing to detect calves
[@ellington_divergent_2020; @decesare_estimating_2012].

To derive a composition survey bias term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
uncollared adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-adjusted recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias term $c$ that is integrated into
the Bayesian population model and simulations (below):

@@

In simulated monitoring scenarios, we assume collars are deployed in January, and
continue to function for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program
$d$ and the ratio of uncollared to collared adults in the recruitment
survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
mortality may be replaced each year ($o=1$), or there may be a 3 year
gap between deployments ($o=4$).

@@

To investigate the effectiveness of various monitoring strategies for
reducing uncertainty about population viability, and to show how this
varies with landscape condition, we begin with a model for simulating
example population dynamics and observations of those dynamics. Demographic parameters for the simulated trajectories are sampled from the regression models that summarize national demographic-disturbance relationships
[@johnson_science_2020] so we call this the "national" model. Together, a set of example trajectories represent alternative hypotheses about the true state of a particular caribou population; these alternative scenarios informed by prior
beliefs are used to assess the expected value of alternative sampling
strategies [@runge_which_2011; @dunham_strategic_2023]. We also use the
simulated trajectories to confirm that the Bayesian model successfully
integrates local data, and that prior predictions meet expectations from the national model. For
reference, all variables and parameters are defined in supplemental
Table A1.

To simulate demographic trajectories for example populations we used a
modified version of Johnson et al's [-@johnson_science_2020] demographic
model described by Dyson et al. [-@dyson_existing_2022]. Given a
population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Realized recruitment rate varies with population density [@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females ($\dot{W}_t$)
and the maximum potential recruitment of female calves per cow ($\dot{X}_t$):

@@

where $w$ is the ratio of observed adult females to
collared animals, $\dot{q}$ is the ratio of young bulls to adult
females, $\dot{u}$ is the probability of misidentifying young bulls as
adult females and vice versa, and $\dot{z}$ is the probability of
missing a calf. We examine sensitivity to these unknown parameters by
selecting values for each simulated example population from plausible
ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and
$\dot{q}$ varies uniformly between 0 and 0.6.

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure that results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period. Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.

@@

@@

To account for sex ratio of calves and additional uncertainty caused by composition survey errors errors (Supplement B) estimated recruitment is adjusted to get expected female
recruits per female: 
\begin{equation}
X_t=cR_t/2; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias term $c$ depends on the known
apparent number of adult females per collared animal $w$, the unknown
ratio of young bulls to adult females $q$, the unknown probability of
misidentifying young bulls as adult females and vice versa $u$, and the
unknown probability of missing a calf $z$ (Supplement B). Prior
uncertainty about the value of the bias term $c$ is
approximated with a Log-normal distribution (Supplement B). Given the
apparent number of adult females per collared animal $w$ in each
monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
the mean and standard deviation of 10,000 samples of $\log{c}$ assuming
the ratio of young bulls to adult females $q$ varies uniformly between 0
and 0.6, the adult misidentification probability $u$ varies uniformly
between 0 and 0.2, and the probability of missing a calf $z$ varies
uniformly between 0 and 0.2.

@@

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression models
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% prediction intervals
from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)).

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is the
probability of correctly predicting the true population status
(increasing or decreasing) [@dunham_strategic_2023]. To calculate this
outcome for each disturbance and monitoring scenario (Fig
\@ref(fig:figDisturbance)) we fit the Bayesian model to simulated
observations from 500 example population trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)), and calculated the percentage of these
500 examples in which the Bayesian posterior expected population growth rate
and the true expected population growth rate are both greater than 0.99, or both less than or equal to 0.99.

@@

EVSI measures expected improvement from a particular sampling strategy
as the difference between the expected value with and without additional
sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\eta_{\alpha,\psi}]] - EV_{current},
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular
example trajectory. In this study, we treat simulated observations from
500 example trajectories for each disturbance and monitoring scenario
$\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the
distribution of possible values of $x$, and assume each of these sample
results is equally probable. $P(\psi|x,\tau)$ is the posterior
probability that the true state of the population is $\psi$, and each
population is assigned a status $\alpha$ informed by the sample
information.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from a national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Figure \@ref(fig:figDiffs)) increase. Reducing uncertainty is more challenging when interannual variation is high (Figure
\@ref(fig:examplePredictions)). Increasing the forecast horizon [@petchey_ecological_2015] does not increase uncertainty about expected population growth (compare 5 and 20 year projections in
Figure \@ref(fig:figDiffs)) in these scenarios.

Our analysis becomes more informative when we examine the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1).
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), but we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are less
precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g 
left side of right panels in Figure \@ref(fig:examplePredictions)). At
lower levels of disturbance, local monitoring can reduce uncertainty, but
there are diminishing returns as the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Acknowledging the possibility of composition survey errors increases uncertainty, but these increases are not large for the set of parameters we investigate
(Supplement F).

@@

@@

@@

Figure 3: Prior (a) and posterior predictions from the Bayesian model for population trajectories with low (b: $v_R=0.23,v_S=0.0435$) and high (c: $v_R=0.46,v_S=0.087$) interannual variation. In this example scenario anthropogenic disturbance increases by 2% per year from 0 to 100%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95% predictive intervals (top row, orange lines and shading) are similar to the means and ranges between the 2.5% and 97.5% quantiles of 3000 simulated survival and recruitment trajectories from the national model (top row, blue lines and shading). Local population data (open triangles) reduces uncertainty about demographic rates, but local data is not required to predict viability (population growth <= 0.99) of highly disturbed populations.

Figure 4: The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).

Figure 5: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).

Figure 6: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort. When disturbance is high EVSI is low, indicating that local sampling does not change predicted status. Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases. Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction. See Figure \@ref(fig:figStatusErrors) for additional scenario details.

>>>>

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g. @flasko_fecal_2017;
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival [e.g. @dalgarno_bboutools_2024; @mcloughlin_declines_2003;
@hervieux_widespread_2013; @eacker_web-based_2019; @lamb_effectiveness_2024]. Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022; @eacker_web-based_2019; @dalgarno_bboutools_2024]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

@@

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, can improve viability projections on changing
landscapes, and can help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach will help to identify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations [@bissonnette_controverses_2020; @boan_climate_2018].

@@

## Overview of methods and notation

To investigate how the accuracy and precision of predictions vary with both landscape condition and monitoring effort we analyze simulated data using a Bayesian model with informative priors and a composition survey bias term (section 2.1). Steps in our analysis procedure include defining anthopogenic disturbance and monitoring scenarios (section 2.2), simulating population dynamics and monitoring (section 2.3) to produce example demographic trajectories for each scenario, and analyzing example demographic trajectories with a Bayesian model that enables integration of (simulated) local demographic data with prior knowledge of demographic-disturbance relationships (section 2.4). Outcomes of interest include expected population growth rate, the probability of correctly assessing population status, and the expected value of local information (section 2.5). We refer to the model used to simulate example trajectories as the "national" model because parameters are from a national analysis of demographic-disturbance relationships [@johnson_science_2020]. We distinguish this national model from the Bayesian model that is used to analyze simulated example demographic trajectories. Where appropriate the national model and the Bayesian model share the same structure, and we have selected notation to highlight the similarities. Bayesian model estimates are distinguished from observed values, estimated expected values, true values, and true expected values using $\hat{\text{ }}$ $\tilde{\text{ }}$ $\dot{\text{ }}$ and $\bar{\text{ }}$ symbols respectively; for example, $S_t$ is the Bayesian estimate of survival in year $t$, $\hat{S}_t$ is observed survival, $\tilde{S}_t$ is the Bayesian estimate of expected survival (without interannual variation), $\dot{S}_t$ is true survival, and $\bar{S_t}$ is true expected survival (without interannual variation). For reference, all variables and parameters are defined in supplemental Table A1.

@@

We focus on analyzing survival of radio-collared females and the ratio of calves to cows in aerial
surveys of groups that include collared females [@nbckc_nbckc_2022],
because these remain the most widely available sources of demographic
information for boreal caribou populations [@johnson_science_2020]. GPS radio collars report animal locations at regular intervals, and cease moving when animals die. Aerial recruitment surveys are usually done in March or April, and radio-collared females are often used to locate groups of animals. Surveyors use visual cues to distinguish calves that have survived their first winter (~ 9 months old) from adult cows and bulls. Surveyed groups also contain uncollared adults so sample size is typically larger for recruitment than for survival estimates. Errors that can bias composition survey results include misidentifying young
bulls as cows or vice versa, and failing to detect calves
[@ellington_divergent_2020; @decesare_estimating_2012].

To derive a composition survey bias term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-adjusted recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to uncollared adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias term $c$ that is integrated into
the Bayesian population model and simulations (below):

@@

In simulated monitoring scenarios, we assume collars are deployed in January, and
continue to function for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program
$d$ and the ratio of observed adult females to
collared animals in the recruitment
survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
mortality may be replaced each year ($o=1$), or there may be a 3 year
gap between deployments ($o=4$).

@@

To investigate the effectiveness of various monitoring strategies, and to show how this
varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories. To simulated example trajectories we use a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al. [-@dyson_existing_2022]. For brevity, we refer to this as the "national" model. For clarity, we include a full description of the model in the following paragraphs. Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population. To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using a Bayesian model (section 2.4). We also use the sets of simulated trajectories to confirm that prior predictions from our Bayesian model are consistent with knowledge of the observed distribution of outcomes for caribou across the country [@johnson_science_2020].

Given a population of post-juvenile females at the beginning of year $t$,
$\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is
binomially distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Realized recruitment rate varies with population density [@lacy_vortex_2017], and the number of juveniles recruiting to the
post-juvenile class at the census $(\dot{J}_{t})$ is a binomially
distributed function of the number of surviving post-juvenile females ($\dot{W}_t$)
and the maximum potential recruitment of female calves per cow ($\dot{X}_t$):

@@

where $w$ is the ratio of observed adult females to
collared animals, $\dot{q}$ is the ratio of young bulls to adult
females, $\dot{u}$ is the probability of misidentifying young bulls as
adult females and vice versa, and $\dot{z}$ is the probability of
missing a calf. We examine sensitivity to these unknown parameters by
selecting values for each simulated example population from plausible
ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and
$\dot{q}$ varies uniformly between 0 and 0.6. The distinction between the true composition bias term $\dot{c}$ and the term $c$ we include in the Bayesian model (eq \@ref(eq:c)) is that the (simulated) truth is known, and the Bayesian model bias is uncertain.

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)).

@@

To ensure that results are comparable across monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period. Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.

@@

To account for sex ratio of calves and additional uncertainty caused by composition survey errors (Supplement B) estimated recruitment is adjusted to get expected female
recruits per female $X_t=cR_t/2$. The composition survey bias term $c$ (eq \@ref(eq:c)) depends on the known
apparent number of adult females per collared animal $w$, the unknown
ratio of young bulls to adult females $q$, the unknown probability of
misidentifying young bulls as adult females and vice versa $u$, and the
unknown probability of missing a calf $z$ (Supplement B). Prior
uncertainty about the value of the bias term $c$ is
approximated with a Log-normal distribution (Supplement B). Given the
apparent number of adult females per collared animal $w$ in each
monitoring scenario (Table \@ref(tab:monitoringScenarios)), we calculate
the mean and standard deviation of 10,000 samples of $\log{c}$ assuming
the ratio of young bulls to adult females $q$ varies uniformly between 0
and 0.6, the adult misidentification probability $u$ varies uniformly
between 0 and 0.2, and the probability of missing a calf $z$ varies
uniformly between 0 and 0.2.

@@

All regression coefficients are assumed to be Gaussian distributed, with
prior means from Johnson et al's (2020) regression models
(Table \@ref(tab:priors)). We calibrated the standard deviations of the
priors so that the the 95% prior predictive intervals for survival and
recruitment from the Bayesian model match the 95% prediction intervals
from Johnson et al's regression model (Supplement C, Figure
\@ref(fig:examplePredictions)). To show the value of informative priors, we also include a weakly informative scenario with prior means of the intercept parameters from Dalgarno et al. [-@dalgarno_bboutools_2024] ($\beta^R_0=-0.3689, \beta^S_0=-0.2126776$), prior means of the slope parameters set to zero, and standard deviations for all parameters (Table \@ref(tab:priors)) multiplied by ten.

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is the
probability of correctly predicting the true population status
(increasing or decreasing) [@dunham_strategic_2023]. To calculate this
outcome for each disturbance and monitoring scenario (Fig
\@ref(fig:figDisturbance)) we simulate observations from 500 example trajectories (e.g. Fig
\@ref(fig:exampleTrajectories)), fit the Bayesian model to each of these, and calculated the percentage of these
500 examples in which the Bayesian posterior expected population growth rate
and the true expected population growth rate are both greater than 0.99, or both less than or equal to 0.99.

@@

EVSI measures expected improvement from a particular monitoring strategy
as the difference between the expected value with and without additional
sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\eta_{\alpha,\psi}]] - EV_{current},
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular
example trajectory. In this study, we treat simulated observations from
500 example trajectories for each disturbance and monitoring scenario
$\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the
distribution of possible values of $x$, and assume each of these sample
results is equally probable. $P(\psi|x,\tau)$ is the posterior
probability that the true state of the population is $\psi$, and each
population is assigned a status $\alpha$ informed by the sample
information.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from a national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). When local data is limited (e.g. 4 years, 15 collars per year) the model with informative priors yields predictions that are consistent with knowledge of observed outcomes across the country, and the model with weakly informative priors does not (Figure
\@ref(fig:examplePredictionsLowEffort)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Figure \@ref(fig:figDiffs)) increase. Interannual variation makes reducing uncertainty challenging, and insufficiently informative priors can yield predictions that are both imprecise and inaccurate even in a case where there is 16 years of local monitoring data (Figure \@ref(fig:examplePredictions)). Increasing the forecast horizon [@petchey_ecological_2015] does not increase uncertainty about expected population growth (compare 5 and 20 year projections in
Figure \@ref(fig:figDiffs)) in these scenarios.

Examining the probability of
correctly assessing population status (Figure \@ref(fig:figStatusErrors))
and the expected value of information (Figures \@ref(fig:figEVSI) and E.1) helps clarify how the need for monitoring varies with landscape condition.
When anthropogenic disturbance is high (right panels of Figures
\@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and E.1)), projections
of population status informed only by priors from the national
demographic-disturbance relationships are accurate, and adding
additional local monitoring data does not improve accuracy.
Interestingly, when the projection period is long and anthropogenic
disturbance is high, we see a negative EVSI (lower right panel of Figure
\@ref(fig:figEVSI)), but we do not see an associated decrease in the
probability of correctly assessing population status (lower right panel
of Figure \@ref(fig:figStatusErrors)). This means that the long-term
projections informed by local monitoring data are less
precise but not less accurate, which can happen if the prior viability predictions are more certain than the posterior predictions (e.g 
left side of right panels in Figure \@ref(fig:examplePredictions)). At
lower levels of disturbance, local monitoring can reduce uncertainty, but
there are diminishing returns as the number of years of monitoring increases (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). Reducing uncertainty is more difficult when interannual variation is high (Figures
\@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)).The shapes of the
relationships between monitoring effort and the probability of correct
status assessment depend on how anthropogenic disturbance is changing
over time, but there are diminishing benefits of increased monitoring
regardless of whether disturbance is increasing or decreasing
(Supplement D). Acknowledging the possibility of composition survey errors increases uncertainty, but these increases are not large for the set of parameters we investigate
(Supplement F).

@@

# Data statement
Code required to create the data, run analyses and reproduce figures in this paper is available at https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper, and in the caribouMetrics R package available on GitHub at https://github.com/LandSciTech/caribouMetrics/tree/EI-paper-submission.

@@

```{r examplePredictionsLowEffort, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamplesLowEffort.png"))
file.rename(here::here("analysis/paper/figs_submit/bayesianModelExamplesLowEffort.pdf"), 
            here::here("analysis/paper/figs_submit/Figure3_bayesianModelExamplesLowEffort.pdf"))
```

@@

Figure 3: Prior (a) and posterior predictions from the Bayesian model for population trajectories with informative (a,b) and weakly informative (c) priors given lower interannual variation ($v_R=0.23,v_S=0.435$) and intensive monitoring. In this example scenario anthropogenic disturbance increases by 2% per year from 0 to 100%. The population is monitored for 16 years with 30 collars per year, and 6 cows per collared cow in the composition surveys. The prior means and 95% predictive intervals (a, orange lines and shading) are similar to the means and ranges between the 2.5% and 97.5% quantiles of 3000 simulated survival and recruitment trajectories from the national model (a, blue lines and shading). Local population data (open triangles) reduces uncertainty about demographic rates (b), though that is not needed to predict viability (population growth <= 0.99) of highly disturbed populations (a). With weakly informative priors the model does not provide accurate or precise predictions of future viability after 16 years of monitoring in a case where anthropogenic disturbance is increasing to 90% (c).

Figure 4: Posterior predictions from the Bayesian model for population trajectories with informative (a) and weakly informative (b) priors given higher interannual variation ($v_R=0.46,v_S=0.087$) and limited monitoring. The population is monitored for 4 years with 15 collars per year, and 3 cows per collared cow in the composition surveys. See Figure 3 for disturbance trends and other details. The limited local data (open triangles) reduces uncertainty (orange shading) about the observed current state of the population, but is not sufficient to reduce future uncertainty. Given limited data the model with informative priors (a) yields predictions (orange bands) that are consistent with knowledge of observed outcomes across the country (blue shading), and the model with weakly informative priors (b) does not.

Figure 5: The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios). Differences decrease as the number of years of monitoring and the number of collars increase. In these examples the composition survey cow multiplier $w$ is 6, collars are renewed each year, and interannual variation is lower ($v_R=0.23,v_S=0.0435$).

Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).

Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort. When disturbance is high EVSI is low, indicating that local monitoring does not change predicted status. Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases. Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction. See Figure \@ref(fig:figStatusErrors) for additional scenario details.
